Predicting Tags for StackOverflow Posts

### 2013-02-11 stackExchange/writeups/iccm-2013/writeup.tex
foo
Former-commit-id: 7421747f5b4edccd4eebe8d1e167c90e0f1c7949 [formerly 59e9d5403ceb470d1edb138840c8900f5a887f9b]
Former-commit-id: f6dc590e4998d77b7d77c3b40e6e3038e9e7e9b7

\begin{abstract}
  A model was built for the StackOverflow dataset that can predict the tags that will be used by the author of the post, given the content of the post.
  The tag-predictor model is a cognitively-inspired Bayesian probabilistic model based on ACT-R's declarative memory retrieval mechanisms.
  Large-scale relational database techniques and efficient sparse matrix data structures were used to implement the model,
  so that the data structures require only 3GB of memory and retrieval requests with a DM dataset of \num{556677795} chunks occur in less than one second on a single processor.
  A logistic regression technique was used to calibrate model parameters and measure performance.
  Model fit was satisfactory ($R_{pseudo}^{2}=.46$, npv = .997, ppv = .70) and model performance in a real-world setting was fairly strong (67\% accuracy when asked to tag one tag per post on average).

@@

A tag-prediction system was built here for the StackOverflow dataset that can solve precisely these sorts of problems.
The StackOverflow dataset was chosen because it is relatively constrained, contains a large number of posts and associated tags, and the data are easy to obtain.
It is constrained because there is a fully enumerated and limited set of tags that an author can use, as opposed to a less-constrained domain such as Twitter.
The site also makes public all posting data on a quarterly basis, which can be easily downloaded and loaded into a relational database for further analysis.

@@

A variation of ACT-R's declarative memory retrieval theory \cite{Anderson2004} was used to model tag retrievals for StackOverflow posts.
The hypothesis is that the tags used for posts are based on a tag's prior use and the strength of association between the tag and the content of words in the title and body of the post.
The strength of association between post words and tags was calculated using 2/3 of the dataset (1M posts).
A logistic regression statistical technique was used to calibrate model parameters and optimize performance.
The weights for each model component were calibrated using a sample of 1,000 posts not contained within the strength of association dataset.
The calibrated model was then tested on a new sample of 1,000 posts not contained within either previous training dataset.
Classification accuracy was used as a metric of model performance (i.e., how many of the model's predicted tags are correct).

@@

The R statistical programming environment was used to build the model and generate model predictions.
This was done by feeding the word co-occurrence and tag occurrence files generated by the database queries into a separate sparse matrix and sparse vector in R.
A hash was computed for each unique word, so that indices for the sparse matrix and vector could be numeric.
The dimensions for the co-occurrence sparse matrix were context word, tag, and the tag occurrence vector was 1 dimensional (tag).

1 million posts (approximately 2/3 of all posts) were fed into R to build the co-occurrences and tag occurrences.
An interface to the model was developed so that one could supply a vector of words, and the model would return a sparse vector of tag activations given that context.
In order to generate a model prediction, tag activations for title and body context were computed separately, and then added to each tag's prior base-level activation.
An offset constant (equal to the average activation of the top 5 model-predicted tags for the post) was then subtracted from the prior value.
The resulting tags were rank sorted, and the one with the highest activation would then be the model's prediction of the most likely tag associated with that post.

@@

These vectors of tag activation, categorization were collected across a set of 1,000 posts.
The set of posts did not include any posts that were used to train the model on word association strength.
So a form of cross validation was used to measure model fit, although the more robust full k-fold cross validation technique was not used (due mostly to resource and time limitations).

@@

\renewcommand{\arraystretch}{1.5}% Wider
\renewcommand{\tabcolsep}{.5mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{The StackOverflow Tag-Prediction Model} 
    \label{sample-table} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN(j,i)}{N_{Row}(j)N_{Col}(i)}$ \\
      Attention Weight	& 	$W_{j} = WE_{j}$ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      Max Entropy & 		$H_{max} = maxH$ \\
      Offset & 			$O = \frac{1}{5}\sum_{i\in top 5}^{ } A_{i}$ \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

@@

\renewcommand{\arraystretch}{1}% Wider
\renewcommand{\tabcolsep}{3mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{Model fit metrics} 
    \label{tab:fits} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Value	\\
      \hline
      McFadden's $R_{pseudo}^{2}$ &	.46 \\
      True Positives &			647 \\
      False Positives &			278 \\
      True Negatives &			\num{796982} \\
      False Negatives &			\num{2235} \\
      Positive Predictive Value &	0.70 \\
      Negative Predictive Value &	0.997 \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

Note that all four coefficients for the run are strong predictors of tag use ($p<.01$), and model fit is satisfactory ($R_{pseudo}^{2}=.46$).
It's also interesting to note that although the attentional weighting term ($W$) was allowed to be calibrated by the regression, it did not stray far from one
(see coefficients for title and body in Table \ref{tab:coeffs}).
Apparently the scaled entropy measure that globally weights each cue allows for total attentional resources ($W$) to remain relatively small (around one) and not stray far from cognitively-plausible defaults (one).

@@

\subsection{Related Work}
Predicting tags for the StackOverflow dataset has been worked on by at least one prior researcher \cite{Kuo2011}.
\citeA{Kuo2011} only looks at the Stack Overflow dataset briefly at the end of the technical report.
However he does mention that the co-occurrence model he used has about a 47\% classification accuracy for predicting StackOverflow tags, when required to predict one tag per post.
The full model used here achieves 67\% classification accuracy when predicting one tag per post on average (height of full model at vertical dotted line in Figure \ref{fig:ROC}.
The increase in model performance is most likely due to FIXME.

@@

However, it is increasingly likely that there is no fundamental reason why vanilla ACT-R (Common Lisp version) cannot be optimized to handle spreading activation calculations with datasets this large.
All that is really needed is a more appropriate data structure to handle the large scale.
This can be seen by comparing the implementation of spreading activation here to the implementation in \citeA{Douglass2010}.
Both implementations handle sji counts greater than \num{1000000}, with declarative memory retrieval times (time to calculate $A_i$'s) around 1 second at this scale.
However, the implementations could not be more different.
The Erlang implementation leverages a semantic network of millions of isolated process chunks and massive concurrency to achieve the scale.
The R implementation here is single threaded and requires only a sparse matrix implementation to achieve the scale.
Since the programming languages are different, it's not likely the case that only a specific programming language can achieve the scale.
And since the parallelism is different (single threaded vs multi threaded), it's not likely the case that achieving the scale can only be done with massively concurrent implementations.
So, it's most likely the case that using an appropriate data structure is all that is needed for ACT-R Common Lisp to handle sji counts in the millions and above.

>>>>

\begin{abstract}
  A model was built for the StackOverflow dataset that can predict the tags that will be used by the author of the post, given the content of the post.
  The model is a cognitively-inspired Bayesian probabilistic model based on ACT-R's declarative memory retrieval mechanisms.
  Large-scale relational database techniques and efficient sparse matrix data structures were used for implementation,
  so that the data structures require only 3GB of memory and retrieval requests with a DM dataset of \num{556677795} chunks occur in less than one second on a single processor.
  A logistic regression technique was used to calibrate model parameters and measure performance.
  Model fit is satisfactory (McFadden's pseudo $R_{}^{2}=.46$, npv = .997, ppv = .70) and model performance on the test dataset is fairly strong (67\% accuracy when asked to tag one tag per post on average).

@@

A tag-prediction system was built here for the StackOverflow dataset that can solve precisely these sorts of problems.
The StackOverflow dataset was chosen because it is relatively constrained, contains a large number of posts and associated tags, and the data are easy to obtain.
It is constrained because there is a fully enumerated and limited set of tags that an author can use, as opposed to a less-constrained domain such as Twitter.
The site also makes public all posting data on a quarterly basis, which can be downloaded and loaded into a local relational database for further analysis.

@@

A variation of ACT-R's declarative memory retrieval theory \cite{Anderson2004} was used to model tag retrievals for StackOverflow posts.
The hypothesis is that the tags used for posts are based on a tag's prior use and the strength of association between the tag and the content of words in the title and body of the post.
The strength of association between post words and tags was calculated using 2/3 of the dataset (one million posts).
A logistic regression statistical technique was used to calibrate model parameters and optimize performance.
The weights for each model component were calibrated using a sample of \num{1000} posts not contained within the strength of association dataset.
The calibrated model was then tested on a new sample of \num{1000} posts not contained within either previous training dataset.
Classification accuracy was used as a metric of model performance (i.e., how many of the model's predicted tags are correct).

@@

The R statistical programming environment was used to build the model and generate model predictions.
This was done by feeding the word co-occurrence and tag occurrence files generated by the database queries into a separate sparse matrix and sparse vector in R.
A hash was computed for each unique word, so that indices for the sparse matrix and vector could be numeric.
The dimensions for the co-occurrence sparse matrix were context word, tag, and the tag occurrence vector was one dimensional (tag).

One million posts (approximately 2/3 of all posts) were fed into R to build the co-occurrences and tag occurrences.
An interface to the model was developed so that one could supply a vector of words, and the model would return a sparse vector of tag activations given that context.
In order to generate a model prediction, tag activations for title and body context were computed separately, and then added to each tag's prior base-level activation.
An offset constant (equal to the average activation of the top 5 model-predicted tags for the post) was then subtracted from the prior value.
The resulting tags were rank sorted, and the one with the highest activation would then be the model's prediction of the most likely tag associated with that post.

@@

These vectors of tag activation, categorization were collected across a set of \num{1000} posts.
The set of posts did not include any posts that were used to train the model on word association strength.
So a form of cross validation was used to measure model fit, although the more robust full k-fold cross validation technique was not used (due mostly to resource and time limitations).

@@

\renewcommand{\arraystretch}{1.5}% Wider
\renewcommand{\tabcolsep}{.5mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{The StackOverflow tag-prediction model} 
    \label{sample-table} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN(j,i)}{N_{Row}(j)N_{Col}(i)}$ \\
      Attention Weight	& 	$W_{j} = WE_{j}$ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      Max Entropy & 		$H_{max} = maxH$ \\
      Offset & 			$O = \frac{1}{5}\sum_{i\in top 5}^{ } A_{i}$ \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

@@

\renewcommand{\arraystretch}{1}% Wider
\renewcommand{\tabcolsep}{3mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{Model fit metrics} 
    \label{tab:fits} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Value	\\
      \hline
      McFadden's Pseudo $R_{}^{2}$ &	.46 \\
      True Positives &			647 \\
      False Positives &			278 \\
      True Negatives &			\num{796982} \\
      False Negatives &			\num{2235} \\
      Positive Predictive Value &	0.70 \\
      Negative Predictive Value &	0.997 \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

Note that all four coefficients for the run are strong predictors of tag use ($p<.01$), and model fit is satisfactory (McFadden's pseudo $R_{}^{2}=.46$).
It's also interesting to note that although the attentional weighting term ($W$) was allowed to be calibrated by the regression, it did not stray far from one
(see coefficients for title and body in Table \ref{tab:coeffs}).
Apparently the scaled entropy measure that globally weights each cue allows for total attentional resources ($W$) to remain relatively small (around one) and not stray far from cognitively-plausible defaults (one).

@@

\subsection{Related Work}
Predicting tags for the StackOverflow dataset has been worked on by at least one prior researcher \cite{Kuo2011}.
\citeA{Kuo2011} only looks at the Stack Overflow dataset briefly at the end of the technical report.
However he does mention that the co-occurrence model he used has about a 47\% classification accuracy for predicting StackOverflow tags when required to predict one tag per post.
The full model used here achieves 67\% classification accuracy when predicting one tag per post on average
(height of full model at vertical dotted line in Figure \ref{fig:ROC}).
The increase in model performance is most likely due to the fact that \citeA{Kuo2011} only used \num{10000} posts to generate the co-occurance matrix ($S_{ji}$).
One million posts (two orders of magnitude higher) was used here.
Also the model here has some additional components: The offset in particular and possibly the body term as well,
as it was unclear from \citeA{Kuo2011} if the author used context from both the body and title of a post or just the title. 
Nonetheless, there are certainly similarities in the approach:
Co-occurance count was used as the core measure of similarity, and both local (strength of association) and global (idf or entropy) measures were used to compute tag likelihood, given context.

@@

However, it is increasingly likely that there is no fundamental reason why vanilla ACT-R (Common Lisp version) cannot be optimized to handle spreading activation calculations with datasets this large.
All that is really needed is a more appropriate data structure to handle the large scale.
This can be seen by comparing the implementation of spreading activation here to the implementation in \citeA{Douglass2010}.
Both implementations handle sji counts greater than \num{1000000}, with declarative memory retrieval times (time to calculate $A_i$'s) around one second at this scale.
However, the implementations could not be more different.
The Erlang implementation leverages a semantic network of millions of isolated process chunks and massive concurrency to achieve the scale.
The R implementation here is single threaded and requires only a sparse matrix implementation to achieve the scale.
Since the programming languages are different, it's not likely the case that only a specific programming language can achieve the scale.
And since the parallelism is different (single threaded vs multi threaded), it's not likely the case that achieving the scale can only be done with massively concurrent implementations.
So, it's most likely the case that using an appropriate data structure is all that is needed for ACT-R Common Lisp to handle sji counts in the millions and above.
